---
title: "Static Maps"
sidebar: false
format:
  pdf: default
editor_options: 
  chunk_output_type: console
author: "Daniel Evans"
  
---




```{r,message=FALSE}
# data
library(tidyverse)
PLACES <- read_csv("Data/PLACES.csv")
PLACES <- PLACES |>
  rename(State = StateDesc)

library(maps)
states <- map_data("state")
head(states)
```


# First Map: Prevalence of Depression Across the United States


```{r}
# First

# summary by state
PLACES_states <- PLACES |>
  mutate(State = tolower(State)) |>
  group_by(State) |>
  summarize(
    continuous = mean(DEPRESSION_CrudePrev), 
    other = mean(SLEEP_CrudePrev)
  )

# join
chimera <- states |>
  left_join(PLACES_states, by = c("region" = "State"))

# plot
chimera |>
  ggplot(aes(x = long, y = lat, group = group)) +
    geom_polygon(aes(fill = continuous), color = "#222200") +
    coord_map() +  
    theme_void() +  
    scale_fill_viridis_c(option = "E") +
    theme(panel.background = element_rect(fill = "beige")) +
    labs(
      title = "Average Prevalence of Depression by State",
      fill = "Mean Prevalence of Depression",
      caption = "Source: CDC's PLACES Census Tract Data 2024 Release\nhttps://data.cdc.gov/500-Cities-Places/PLACES-Census-Tract-Data-GIS-Friendly-Format-2024-/yjkw-uj5s"
    )
```


**Alt-Text Description:**
This is a choropleth map or plot of the United States. Each state is drawn out and colored according to its average prevalence of depression, which ranges from 16 to 29 percent. The states with the lowest rates of depression appear to be Nebraska and New Jersey, while the highest prevalence of depression occurring in Tennessee and West Virginia. In general, the pattern of prevalence of depression does not seem to be in a relationship with political trends, rural vs urban dynamics, nor coastal vs mainland geographical locations. 





# Second Map: Prevalence of Smoking Across the United States


```{r}
# First
AVG <- mean(PLACES_states$other)

chimera <- chimera |>
  mutate(categorical = ifelse(other>AVG, "More Than 35.8%", "Less Than 35.8%"))


# plot
chimera |>
  ggplot(aes(x = long, y = lat, group = group)) +
    geom_polygon(aes(fill = fct(categorical)), color = "#353535") +
    coord_map() +  
    theme_void() +  
    scale_fill_viridis_d(option = "F") +
    theme(panel.background = element_rect(fill = "lavender")) +
    labs(
      title = "U.S. States With Higher Than Average Sleep Deprivation",
      fill = "Prevalence of Sleep Deprivation",
      caption = "Source: CDC's PLACES Census Tract Data 2024 Release\nhttps://data.cdc.gov/500-Cities-Places/PLACES-Census-Tract-Data-GIS-Friendly-Format-2024-/yjkw-uj5s"
    )
```


**Standard Description:**
BLABLABLABLABLA











